Rheology of Cubic Blue Phases

### 2013-03-11 trunk/papers/bprheo/bprheo.tex
More detail on method in section 2
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1938 40876441-3496-0410-a613-a2fca35b2ae2

In what follows, $x$, $y$ and $z$ denote respectively the velocity, velocity
gradient and vorticity direction -- therefore $\Pi_{xy}$ is the shear stress.

The system of coupled partial differential equations Eqs.~\ref{op-eom}
and~\ref{NSE} is solved by means of a
hybrid scheme~\cite{Marenduzzo:2007}, which combines different numerical schemes
and provides a way to address the multiscale nature of the problem.
More details on the algorithm and the boundary conditions can
be found in~\cite{Denniston:2001, Denniston:2004, Henrich:2012a}.

In the following we report results of simulations of the bulk flow behaviour of the cubic blue 
phases BPI and BPII.
Typical runtimes for system size $L_x\times L_y\times L_z=128^3$ on 512 processes were in the region of 18 to 24 hours.  
The timestep and lattice spacing in lattice Boltzmann units (LBU) can be mapped
approximately to $\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI units, respectively. The LB unit of stress
is equal to about $10^8$~Pa. Further details about the conversion 
from LBU to SI units can be found in ~\cite{Henrich:2011a,Henrich:2010b}.

@@

For typical simulations reported in this work, 
we chose as initial conditions thermodynamic states that are 
well inside the equilibrium region of the individual blue phase, and far away 
from the cholesteric-isotropic transition. Thus, temperature and chirality were 
$\tau=-0.5, \kappa=1.0$ in case of BPI and $\tau=-0.5, \kappa=2.0$ for BPII, respectively.
For these parameters the total free energy density $f$ remained always negative at all flow rates
simulated. 
Since by Eq.~\ref{FE} $f=0$ for an isotropic phase with ${\mathbf Q}\equiv 0$, this means that our system, which is never far from equilibrium locally even under flow, always remains in a liquid crystalline state. 
We performed also selected simulations on metastable states at higher and lower temperatures, 
and at different chiralities, but did not find any significant differences in the 
general flow behaviour from that described below.
The only quantitative difference we found was that, for thermodynamic states that are closer to the phase boundary than the one we focused on (and describe below), the critical shear
rate at which the disclination 
network broke up into chiral nematic states was lower. This is expected as these states have on average higher free energy densities and smaller order parameters than those we focus on in what follows, and cannot therefore withstand the same external forces before breaking down.

@@

Simple shear flow was imposed by means of a Lees-Edwards boundary condition ~\cite{Wagner:2002} with
the top (bottom) part of the system flowing in the positive (negative) $x$-direction and the 
velocity gradient along the $y$-direction.
The shear rates were varied over more than two orders of magnitude from about 
$\gd=2.44\times \e{-6}$ to $1.875\times\e{-3}$ LBU.
For clarity we classify various flow regimes, two in the case of BPII and three in the 
case of BPI. 
They are BPII-1 ($\gd \lesssim 3.125\e{-4}$, corresponding to ${\it Er} \lesssim 2.67$), 
BPII-2 ($4.687\lesssim \gd\lesssim 6.25\e{-4}$; $4\lesssim {\it Er} \lesssim 5.34$), ),
BPI-1 ($\gd \lesssim 1.95\e{-5}$; ${\it Er} \lesssim 0.33$), ), BPI-2 ($3.91\e{-5}\lesssim \gd \lesssim 4.687\e{-4}$; $0.67\lesssim {\it Er} \lesssim 8$) and BPI-3 ($6.25\e{-4}\lesssim \gd\lesssim 7.813\e{-4}$; $10.67 \lesssim {\it Er} \lesssim 13.33$).
Beyond these regimes, and prior to the ultimate breakdown into an isotropic phase, 
both BPI and BPII form common configurations regardless 
of the initial state. These are a cholesteric helix with helical axis 
oriented along the direction of the velocity gradient (CHG), 
and a flow-aligned nematic state (FAN).  
As is standard \cite{Henrich:2010b,Henrich:2012b} disclination lines are represented by plotting an isosurface of the scalar order parameter $q_s$. 
Typical choices are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.
(Note that $q_s$ is small but non-zero at the disclination core. The director is undefined there because the largest and second largest eigenvalues of ${\mathbf Q}$ coincide.)

@@

In this Appendix we give some more details on our simulations,
including the importance and effect of boundary conditions.

>>>>

The system of coupled partial differential equations~\ref{op-eom}
and~\ref{NSE} is solved by means of a hybrid method~\cite{Marenduzzo:2007}
which uses a combination of lattice Boltzmann and finite difference schemes.
(This is in contrast with some earlier methods using solely
LB~\cite{Denniston:2001, Denniston:2004}.)
The Navier-Stokes
equation is solved via the lattice Boltzmann approach, using a standard
three-dimensional model with 19 discrete velocities (D3Q19).
A regular lattice with spacing $\Delta x = \Delta y = \Delta z = 1$ is
used and the time step is $\Delta t = 1$ in lattice units.
Coupling to the thermodynamic sector is via a
local body force computed as the divergence of the thermodynamic
stress Eq.~\ref{Pi}; the resulting velocity field is used in the computation
of the time evolution of $\mathbf{Q}$ via a standard finite difference
method using the same grid and the same time step as the LB. The system
is periodic in all three coordinate directions. Sliding
boundary conditions, or Lees-Edwards planes, are explicitly implemented
in both hydrodynamic and thermodynamic sectors to impose shear on the
system. On
crossing a Lees-Edwards boundary a Galilean transformation is applied
with a velocity increment which is fixed in time (and limited in size by the
low Mach number constraint of LB). An appropriate transformation
of LB distributions which propagate across a boundary~\cite{Wagner:2002}
must be applied at each time step, and appropriate
adjustment to $W_{\alpha\beta}$ is required to compute cross-plane
gradients of the velocity field used in the update to ${\mathbf Q}$.
In both cases interpolation of the
relevant quantities is required to cope with the relative displacement
of neighbouring lattice sites separated by a sliding plane (the
displacement may be a fraction of a lattice unit at any given time
step). The use of multiple sliding planes equally spaced in a single
system allows
the overall shear rate to be maintained indefinitely as the system
becomes larger
in the velocity gradient direction (in contrast with the use of solid
walls to impose shear). This hybrid method with Lees-Edwards planes
has been used successfully to study e.g., smectics~\cite{Henrich:2012a}.

In the following we report results of simulations of the bulk flow behaviour of the cubic blue 
phases BPI and BPII.
Typical runtimes for system size $L_x\times L_y\times L_z=128^3$ on 512 processes were in the region of 18 to 24 hours.  
The timestep and lattice spacing in lattice Boltzmann units (LBU) can be mapped
approximately to $\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI units, respectively. The LB unit of stress
is equal to about $10^8$~Pa. Further details about the conversion 
from LBU to SI units can be found elsewhere~\cite{Henrich:2011a,Henrich:2010b}.
In what follows, $x$, $y$ and $z$ denote respectively the velocity, velocity
gradient and vorticity direction; $\Pi_{xy}$ is therefore the shear stress.

@@

For typical simulations reported in this work, 
we chose as initial conditions thermodynamic states that are 
well inside the equilibrium region of the individual blue phase, and far away 
from the cholesteric-isotropic transition. Thus, temperature and chirality were 
$\tau=-0.5, \kappa=1.0$ in case of BPI and $\tau=-0.5, \kappa=2.0$ for BPII, respectively.
For these parameters the total free energy density $f$ remained always negative at all flow rates
simulated. 
Since by Eq.~\ref{FE} $f=0$ for an isotropic phase with ${\mathbf Q}\equiv 0$, this means that our system, which is never far from equilibrium locally even under flow, always remains in a liquid crystalline state. 
We also performed selected simulations on metastable states at higher and lower temperatures, 
and at different chiralities, but did not find any significant differences in the 
general flow behaviour from that described below.
The only quantitative difference we found was that, for thermodynamic states that are closer to the phase boundary than the one we focused on (and describe below), the critical shear
rate at which the disclination 
network broke up into chiral nematic states was lower. This is expected as these states have on average higher free energy densities and smaller order parameters than those we focus on in what follows, and cannot therefore withstand the same external forces before breaking down.

@@

Simple shear flow was imposed by means of the Lees-Edwards boundary
conditions with
the top (bottom) part of the system flowing in the positive (negative) $x$-direction and the 
velocity gradient along the $y$-direction.
The shear rates were varied over more than two orders of magnitude from about 
$\gd=2.44\times \e{-6}$ to $1.875\times\e{-3}$ LBU.
For clarity we classify various flow regimes, two in the case of BPII and three in the 
case of BPI. 
They are BPII-1 ($\gd \lesssim 3.125\e{-4}$, corresponding to ${\it Er} \lesssim 2.67$), 
BPII-2 ($4.687\lesssim \gd\lesssim 6.25\e{-4}$; $4\lesssim {\it Er} \lesssim 5.34$), ),
BPI-1 ($\gd \lesssim 1.95\e{-5}$; ${\it Er} \lesssim 0.33$), ), BPI-2 ($3.91\e{-5}\lesssim \gd \lesssim 4.687\e{-4}$; $0.67\lesssim {\it Er} \lesssim 8$) and BPI-3 ($6.25\e{-4}\lesssim \gd\lesssim 7.813\e{-4}$; $10.67 \lesssim {\it Er} \lesssim 13.33$).
Beyond these regimes, and prior to the ultimate breakdown into an isotropic phase, 
both BPI and BPII form common configurations regardless 
of the initial state. These are a cholesteric helix with helical axis 
oriented along the direction of the velocity gradient (CHG), 
and a flow-aligned nematic state (FAN).  
As is standard \cite{Henrich:2010b,Henrich:2012b} disclination lines are represented by plotting an isosurface of the scalar order parameter $q_s$. 
Typical choices are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.
(Note that $q_s$ is small but non-zero at the disclination core. The director is undefined there because the largest and second largest eigenvalues of ${\mathbf Q}$ coincide.)

@@

In this Appendix we give some more details on our simulations,
including the relationship between boundary conditions and
permeation flow.
